 

### 2021-06-07 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

**Canopy foliage acts as the primary physical barrier between the atmosphere and the forest floor, buffering multiple aspects of the understory conditions. It is critical in influencing -- and is influenced by -- the vertical biophysical gradient (Fig. `r fig_NEON_vertical`).**
 The distribution of this foliage is structured in correspondence with the distribution of tree sizes within a forest, where leaf area density ($LAD$) ie. leaf area per unit volume, increases with tree-height eventually forming the canopy layer.

LAD along the vertical gradient is heterogeneous, with common patterns that vary across forests. Dense canopy forests composed of broadleaf trees can have high $LAD$ in the upper canopy layers, as well as in the understory for eg. in tropical forests [@ashtonComparisonsStructureMixed1992; @koikeCanopyStructureTropical1993] whereas in temperate forests, although LAD is generally more concentrated in the upper-canopy, undualting patterns are formed by conifers and top vs. bottom heavy foliage trees and sometimes by dense understories [@parkerVerticalProfileCanopy1989; @terborghVerticalComponentPlant1985]. In open canopy forests, such as needle-leaf boreal forests, LAD is predominantly accumulated in the lower canopy layers [@lawLeafAreaDistribution2001; @baldocchiSeasonalVariationEnergy1997; @bonan_ecological_2016].SOil moisture conditions, topography, gap formations followed by disturbance, all alter foliage patterns and further influence the formation of relatively closed vs. open canopy forests [e.g., @parkerCanopyLightTransmittance2002 ; @almeidaContrastingFireDamage2016 ; @starkReframingTropicalSavannization2020; @bonan_ecological_2016]. In addition, seasonally dry and wet conditions, deciduousness and phenology contribute to the spatial loss and gain of leaves (Parker & Tibbs 2004; @smithSeasonalDroughtrelatedChanges2019 ; @parkerConsequencesEnvironmentalHeterogeneity2019). In mixed deciduous- evergreen broadleaf forests, the fraction of deciduous trees tends to increase across tree size classes (@conditQuantifyingDeciduousnessTropical2000c, Meakem, *temperate REFS*). In this review, we focus on growing season conditions unless otherwise noted.

@@

Across forest species, widely, show fundamental trait adaptation to light gradients similar to within-canopy traits [@rozendaalPlasticityLeafTraits2006; @thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012]. However, between multi-species canopy and understory, trait plasticity to similar biophysical gradients can vary several-folds based on inherent specie-specific traits. Spatio-temporal gradients offer another axis for trait variation, as they are heterogenous in time and space [Ishii and Asano, 2007]. Across ontogeny of the same species in canopy and understory, traits such as LMA, photosynthetic capacity [per area] and $gs$ increase with height, separate from light [@thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012; @cavender-baresChangesDroughtResponse2000]. Therefore $T_{leaf}$ varies across speci-specific traits, where species with smaller leaves and larger $gs$ will be less vulnerable to temperature increase [@fauset_differences_2018; @songTreeSurfaceTemperature2020].

@@

Exposed-sun-leaves, though equipped with thermoregulatory traits are largely constrained by $gs$ than shade-leaves [@sanchesDifferentialLeafTraits2010; @hernandezSimilarTemperatureDependence2020] because although evapotranspiration and $gs$ peak during the day, during the midday and afternoon $gs$ declines more steeply in sun-leaves resulting in increased $T_{leaf}$ [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004; @kochDiurnalPatternsLeaf1994b; @kosugiSeasonalFluctuationsTemperature2006].

>>>>

**Canopy foliage acts as the primary physical barrier between the atmosphere and the forest floor, buffering multiple aspects of the understory conditions. It is critical in influencing -- and is influenced by -- the vertical biophysical gradient (Fig. `r fig_NEON_vertical`).**
 The distribution of this foliage is structured in correspondence with the distribution of tree sizes, and successional stages within a forest, where leaf area density ($LAD$) ie. leaf area per unit volume, increases with tree-height eventually forming the canopy layer.

LAD along the vertical gradient is heterogeneous, with common patterns that vary across forests. Dense canopy forests composed of broadleaf trees can have high $LAD$ in the upper canopy layers, as well as in the understory for eg. in tropical forests [@ashtonComparisonsStructureMixed1992; @koikeCanopyStructureTropical1993] whereas in temperate forests, although LAD is generally more concentrated in the upper-canopy, undualting patterns are formed by conifers and top vs. bottom heavy foliage trees and sometimes by dense understories [@parkerVerticalProfileCanopy1989; @terborghVerticalComponentPlant1985; @ishiiExploringRelationshipsCanopy2004a]. In open canopy forests, such as needle-leaf forests, LAD is predominantly accumulated in the lower canopy layers [@lawLeafAreaDistribution2001; @baldocchiSeasonalVariationEnergy1997; @bonan_ecological_2016; @ishiiExploringRelationshipsCanopy2004a].Soil moisture conditions, topography, gap formations followed by disturbance, all alter foliage patterns and further influence the formation of relatively closed vs. open canopy forests [e.g., @parkerCanopyLightTransmittance2002 ; @almeidaContrastingFireDamage2016 ; @starkReframingTropicalSavannization2020; @bonan_ecological_2016]. In addition, seasonally dry and wet conditions, deciduousness and phenology contribute to the spatial loss and gain of leaves (Parker & Tibbs 2004; @smithSeasonalDroughtrelatedChanges2019 ; @parkerConsequencesEnvironmentalHeterogeneity2019). In mixed deciduous- evergreen broadleaf forests, the fraction of deciduous trees tends to increase across tree size classes (@conditQuantifyingDeciduousnessTropical2000c, Meakem, *temperate REFS*). In this review, we focus on growing season conditions unless otherwise noted.

@@

Across forest species, widely, show fundamental trait adaptation to light gradients similar to within-canopy traits [@rozendaalPlasticityLeafTraits2006; @thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012]. However, between multi-species canopy and understory, trait plasticity to similar biophysical gradients can vary several-folds based on inherent specie-specific traits. Spatio-temporal gradients, offer another axis for trait variation, as they are heterogenous in time and space, with different levels of trait plasticity [Ishii and Asano, 2007; @ishiiExploringRelationshipsCanopy2004a]. Across ontogeny of the same species in canopy and understory, traits such as LMA, photosynthetic capacity [per area] and $gs$ increase with height, separate from light [@thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012; @cavender-baresChangesDroughtResponse2000]. Therefore $T_{leaf}$ varies across speci-specific traits, where species with smaller leaves and larger $gs$ will be less vulnerable to temperature increase [@fauset_differences_2018; @songTreeSurfaceTemperature2020].

@@

Exposed-sun-leaves, though equipped with thermoregulatory traits are largely constrained by $gs$ than shade-leaves [@sanchesDifferentialLeafTraits2010; @hernandezSimilarTemperatureDependence2020] where although evapotranspiration and $gs$ peak during the day, during the midday and afternoon $gs$ declines more steeply in sun-leaves resulting in increased $T_{leaf}$ [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004; @kochDiurnalPatternsLeaf1994b; @kosugiSeasonalFluctuationsTemperature2006]. This diurnal pattern and resulting $A_{max}$ differs between species, such as early- and late-successional [@ishiiExploringRelationshipsCanopy2004a], and between gymnosperms and angiosperms [@bonan_ecological_2016].
